A general model for the evolution of nuptial gift-giving

### 2024-06-03 ms_refs_fixed.tex
Edit the supporting information to update based on reviewer comments and some minor corrections given the IBM

The rates at which males encounter females \(R_{\mathrm{f},\mathrm{m}}\)
and females encounter males with nuptial gifts
\(R_{\mathrm{m_{G}}, \mathrm{f}}\) are calculated directly from the IBM,
thereby modelling how these rates might be estimated from empirical
data, so the male threshold for increasing fitness by searching is,

@@

Consequently, the IBM and the analytical model differ slightly (e.g.,
time is discrete in the IBM but continuous the analytical model, and in
the IBM a fitness increment is applied to the focal female in the form
of birthrate increase rather than offspring; see Supporting Information
S1 for details). But the predicted thresholds are theoretically
equivalent and yield predictions that are qualitatively the same (Figure
3).

\begin{figure}
\centering
\includegraphics{ms_refs_fixed_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{The coevolution of male search and female choosiness as a
function of nuptial gift search time (\(\alpha\)). Points show where the
lower 95\% confidence interval of female choosiness (red) and male
search (blue) exceeds zero, indicating evolution of choosiness or
nuptial gift search. Each point includes data from 3200 replicate
simulations with identical starting conditions. Red and blue lines show
thresholds above which the mathematical model predicts that females
should be choosy and males should search, respectively (in agreement
with Figure 2). Minor deviations between the analytical model and the
simulation results are expected due to the finite population size and
the substantial stochasticity inherent to the simulation model (for
details, see Supporting Information S1). Approximately 3000 interactions
occur between individuals in each time step, potentially resulting in a
mating interaction. The number of individuals in the population remained
at or near carrying capacity of \(K = 1000\). Expected female processing
time was set to \(T_{\mathrm{f}}=2\) time steps, and \(\gamma\) and
\(\alpha\) values in the range {[}0.0, 2.0{]} and {[}0.1, 2.0{]},
respectively, were used.}
\end{figure}

Male thresholds \(\gamma_{\mathrm{m, IBM}}\) given by ineq. 4 accurately
predict the evolution of searching in the IBM across \(\alpha\) values,
and the female threshold \(\gamma_{\mathrm{f, IBM}}\) (ineq. 5)
accurately predicts the evolution of female choice (Figure 3). In other
words, IBM simulations demonstrate that nuptial gift search in males,
and choosiness in females, will evolve from an ancestral state of no
searching and no choosiness in similar parameter space (Figure 3) as
predicted by the analytical model (Figure 2). We further ran the IBM
with realistic values of \(\gamma\), estimated using data from the
\emph{P. mirabilis} system wherein choosiness among females, and nuptial
gift search among males, occur (Figure 4). We found that our IBM
predicts both the evolution of choosiness and nuptial gift searching
observed in the \emph{P. mirabilis} system.

\begin{figure}
\centering
\includegraphics{ms_refs_fixed_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{The joint evolution of male search and female choosiness using
a nuptial gift fitness increment (\(\gamma\)) that was estimated from
experimental data for the species \emph{P. mirabilis} (mean
\(\gamma = 3.29 \pm\) SE via propagation of error for estimates of
nuptial feeding a non-nuptial feeding groups: {[}1.47; 8.32{]}). Points
show where the 95\% confidence interval exceeds 0 for female choosiness
(red) and male search (blue). Each point includes data from 100
replicate simulations with identical starting conditions. The red line
shows the threshold above which females should be choosy and the blue
line shows the threshold above which males should search. As predicted
by the analytical model, both male search and female choosiness evolved
for a range of \(\gamma\) values around the empirical estimates. This
occurs because these values result in a \(\gamma\) above the fitness
threshold necessary for selection to favour male search of nuptial gifts
(blue line) and female choosiness (red line). Approximately 3000
interactions occur between individuals in each time step, potentially
resulting in a mating interaction. The number of individuals in the
population remained at or near carrying capacity of \(K = 1000\). Female
processing time was set to \(T_{\mathrm{f}}=2\) time steps, and
\(\gamma\) and \(\alpha\) values in the {[}0.2, 2.0{]}, respectively,
were used. The parameter \(\gamma\) was estimated as relative increase
in offspring production such that \(\gamma\) is the factor by which
fitness increases (relative to the baseline fitness) given a nuptial
gift (See Supporting Information S5).}
\end{figure}

@@

Importantly, we show that the threshold nuptial gift value at which
females are favoured to express choosiness for nuptial gifts is rarely
equivalent to the threshold value at which males are favoured to search
for nuptial gifts, potentially leading to sexual conflict
(\protect\hyperlink{ref-Arnqvist2005a}{Arnqvist \& Rowe, 2005};
\protect\hyperlink{ref-Oliveira2008}{Oliveira \emph{et al.}, 2008}).
Here, we are defining sexual conflict as occurring when interactions
between sexes result in situation where both sexes cannot achieve an
optimal outcome simultaneously
(\protect\hyperlink{ref-Parker2006}{Parker, 2006}). As an example,
sizable areas of parameter space exists wherein the female optimum would
be to exhibit preference for (and receive) nuptial gifts, while the male
optimum is to not search for (and give) nuptial gifts (see Figure 2,
Zone C). In many systems, ecological variables such as search time
required to find a nuptial gift will likely depend on prey abundance,
which can vary substantially with time in some species with nuptial
gift-giving (\protect\hyperlink{ref-Ghislandi2018}{Ghislandi \emph{et
al.}, 2018}). Since several ecological variables likely affect the value
of these thresholds, our results can be seen as providing some
formalised description of why nuptial gift-giving only occurs in some
but not all systems.

At first, the analytical model seems to suggests that nuptial gifts must
cause a very high fitness increase (approximately 25\%) before male
search and female choosiness is favoured by selection (Figure 2).
Similarly the IBM model seems to suggest that a fitness benefit of
approximately 50\% is required (see Figure 3). However, it is important
to point out that these thresholds depend on multiple parameters. For
example, if female processing time (\(T_{\mathrm{f}}\)) is high, the
female threshold for choosiness with respect to \(\gamma\) drops such
that male search and female choosiness are favoured at lower \(\gamma\)
(see Supporting formation S1). If \(T_{f}\) is sufficiently high, then
an initially rare gift-giving trait might therefore be favoured by
selection even if the fitness benefit of a nuptial gift is low. The
effect that nuptial gifts have on fitness might vary across systems, or
even populations. Effects on female fecundity have been estimated in
crickets, fireflies, butterflies, and spiders, but these estimates vary
considerably between species suggesting a large positive effect to no
effect at all (\protect\hyperlink{ref-Bergstrom2002}{Bergström \&
Wiklund, 2002}; \protect\hyperlink{ref-Rooney2002}{Rooney \& Lewis,
2002}; \protect\hyperlink{ref-Maxwell2018}{Maxwell \& Prokop, 2018};
\protect\hyperlink{ref-Gao2019}{Gao \emph{et al.}, 2019}).

When modelling nuptial gift evolution, the challenge is to construct a
modelling framework that captures the frequency-dependent selection
between male nuptial gift-giving and female preference for nuptial
gifts, and we do this using a time-in, time-out model. Recent studies
have modelled some frequency-dependent aspect of nuptial gift giving
using evolutionary game theory
(\protect\hyperlink{ref-MaynardSmith1982}{Maynard Smith, 1982};
\protect\hyperlink{ref-Vincent2005}{Vincent \& Brown, 2005}). Two such
studies formulated a quantitative genetics model to study evolutionarily
stable nuptial gift sizes in populations where the female propensity to
remate was evolving (\protect\hyperlink{ref-Kamimura2021}{Kamimura
\emph{et al.}, 2021}; \protect\hyperlink{ref-Iwasa2022}{Iwasa \&
Yamaguchi, 2022}). The results obtained in these studies complement our
results by giving equilibrium solutions to the evolutionary stable
nuptial gift size, whereas we determine the general conditions under
which nuptial gift-giving will evolve as given by the inequalities we
derive.

@@

@@

Our model assumes something akin to sequential polyandry. That is, a
system wherein female mating and reproduction with multiple males occurs
in sequence, rather than multiple matings occurring before reproduction.
In some systems with nuptial gift-giving, females have been documented
to mate multiple times before reproduction occurs, including the genus
of bark lice \emph{Neotrogla}
(\protect\hyperlink{ref-Kamimura2021}{Kamimura \emph{et al.}, 2021}),
and even our example system of \emph{P. mirabilis} where females will
sometimes engage in multiple mating before reproducing, especially if
starved because multiple mating may result in more nuptial gifts
(\protect\hyperlink{ref-Toft2015}{Toft \& Albo, 2015};
\protect\hyperlink{ref-Matzke2022}{Matzke \emph{et al.}, 2022}). It is
unclear what effect (if any) assuming non-sequential polyandry would
have on the threshold we derive. Under non-sequential polyandry, a
viable strategy for females might be to accept any male (with or without
gift) for fertilisation assurance, then exhibit a preference for nuptial
gifts. This might make choosiness less costly since it would entail less
of an opportunity cost to be choosy, and this could potentially make
female preference for nuptial gifts more likely to evolve. Our model
also assumes that males search in time out, rather than contribute to
parental care, which is likely to be accurate for most systems but not
all. Expanding the model to explore these possibilities would be a
worthwhile goal for future research.

@@

Here we first present more detail for the derivation of fitness
threshold values \(\gamma_{\mathrm{m}}\) and \(\gamma_{\mathrm{f}}\). We
then present full details for IBM simulations (see Supporting
Information S1). Code for simulations is available on GitHub (see ``Data
availability'').

@@

After mating, females must spend some time processing offspring
(\(T_{\mathrm{f}}\)). Male time to replenish sperm is assumed to be
negligible, but males can spend time out of the mating pool to search
for a nuptial gift (\(T_{\mathrm{m}}\)). When males return from
time-out, they encounter females with some probability that is a
function of the rate at which an individual encounters conspecifics
(\(R\)) and the sex ratio (\(\beta\); males/females). Mortality occurs
for females and males in (\(\mu_{\mathrm{i,f}}\),
\(\mu_{\mathrm{i,m}}\)) and out (\(\mu_{\mathrm{o,f}}\),
\(\mu_{\mathrm{o,m}}\)) of the mating pool. Following Kokko \& Ots
(\protect\hyperlink{ref-Kokko2006}{2006}), we assume
\(m_{\mathrm{i,f}} = m_{\mathrm{o,f}} = 1\) and
\(m_{\mathrm{i,m}} = m_{\mathrm{o,m}} = 1\). While this choice is
arbitrary, we conducted a sensitivity analysis which shows that the
mortality parameters have no influence on the propensity for male search
and female choice to evolve (Supporting Information S2). First, we
describe the fitness consequences of male search time for a nuptial
gift. We then describe the fitness consequences of female choice to
accept or reject males based on their provision of a nuptial gift.

@@

During time-out, males have the opportunity to search for a nuptial
gift. Males can adopt one of two strategies; either search or do not
search for a nuptial gift. Males with the former strategy continue to
search until they find a nuptial gift, while males that do not search
will immediately re-enter the mating pool. In this case, time searching
for a nuptial gift will come at the cost of mating opportunities but
might increase offspring fitness offspring. We therefore need to model
the expected length of time \(E[T_{\mathrm{m}}]\) spent outside of the
mating pool for males that search for nuptial gifts, which is simply
\(\alpha\). Note that we can integrate search time \(t\) over the rate
at which nuptial gifts are encountered (\(\exp(-1/\alpha)\)) to show
\(E[T_{\mathrm{m}}] = \alpha\),

@@

The rate at which a focal male that searches for a nuptial gift
increases his fitness is therefore the fitness of his offspring
\((1 + \gamma)\) divided by expected time spent searching for a nuptial
gift (\(\alpha\)) plus time spent in the mating pool, \((\beta + 1)/R\)
(recall that females produce \(\lambda\) offspring),

\[W_{m,G} = \lambda \frac{1 + \gamma}{\alpha + \left( \frac{\beta + 1}{R} \right)}.\]

In contrast, a male that does not search for a nuptial gift has
offspring with lower fitness, but spends less time outside of the mating
pool,

\[W_{m,L} = \lambda \frac{1}{\left(\frac{\beta+1}{R} \right)} = \lambda \frac{R}{\beta + 1}.\]

We can then determine the conditions for which
\(W_{\mathrm{m},G} > W_{\mathrm{m},L}\), isolating \(\gamma\) to find
how large of a fitness benefit must be provided by the nuptial gift to
make the search cost worthwhile, which simplifies to ineq. 2. When this
inequality holds, males are favoured to search until they find a nuptial
gift, which would result in an average search time of \(\alpha\). When
the male trait is continuous (i.e., males search for time period
\(T_{\mathrm{m}}\)), it can be shown that the same threshold can be
reached by evaluating the partial derivative of the male fitness
function (Supporting Information S3). Hence, the thresholds are
consistent under different assumptions concerning male searching
strategy. Selection will cause males to search for nuptial gifts if the
fitness increase to offspring exceeds the product of search time and
female encounter rate.

@@

\[R_{f,G} = R \left(\frac{\beta}{\beta + 1}\right)\left(1 - e^{-\frac{1}{\alpha}T_{m}}\right).\]

@@

\[R_{f,L} = R \left(\frac{\beta}{\beta + 1}\right)\left(e^{-\frac{1}{\alpha}T_{m}}\right).\]

@@

If \(R_{\mathrm{f},G}\) is sufficiently high and \(R_{\mathrm{f},L}\) is
sufficiently low, then finding a male with a gift will be easier than
finding a male without one. Also, the expected time spent in the mating
pool before a focal female encounters a male with a gift will be
\(1/R_{\mathrm{f},G}\), while the expected time spent in the mating pool
before a focal female encounters any male will be
\(1 / (R_{\mathrm{f},G} + R_{\mathrm{f},L})\). Finally, the rates at
which a female encounters males with and without a gift,
\(R_{\mathrm{f},G}\) and \(R_{\mathrm{f}, L}\), are different from the
probabilities that a male encounter has or does not have a gift. The
rate of encounter is no longer relevant in this case because we are
assuming that an encounter has occurred. Hence, the probability of an
encountered male having a gift is simply,

\[G = \frac{1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}}{\left(1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}\right) + e^{-\frac{1}{\alpha}T_{\mathrm{m}}}} = 1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@

\[L = e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@

\begin{equation}
W_{\mathrm{f}, G} = \lambda \frac{1 + \gamma}{T_{\mathrm{f}} + \frac{1}{R_{\mathrm{f},G}}}.
\end{equation}

The top of the right-hand side of Eq. 6 gives the fitness increase, and
the bottom gives the total time it takes to obtain this fitness. The
\(R_{\mathrm{f},G}\) is inverted because it represents the expected time
to encountering a male with a gift. We can expand Eq. 6,

\[W_{\mathrm{f},G} = \lambda \frac{1 + \gamma}{T_{f} + \frac{1}{R \left(\frac{\beta}{\beta + 1}\right)\left(1 - e^{-\frac{1}{\alpha}T_{m}}\right)}}.\]

@@

\[W_{\mathrm{f},l} = \lambda \frac{\left(1 + \gamma\right)\left(1 - e^{-\frac{1}{\alpha}T_{m}}\right) + e^{-\frac{1}{\alpha}T_{m}}}{T_{f} + \frac{1}{R \left(\frac{\beta}{\beta + 1}\right)}}.\]

We then evaluate the conditions under which
\(W_{\mathrm{f}, G} > W_{\mathrm{f}, L}\). We isolate \(\gamma\) to
determine how much offspring fitness must be increase to make choosiness
beneficial (\(\gamma_{\mathrm{f}}\)),

@@

We assume that the sex ratio is equal upon maturation. Given this, Kokko
and Monaghan (\protect\hyperlink{ref-Kokko2001}{Kokko \& Monaghan,
2001}) show that the operational sex ratio depends on the probability of
finding an individual in `time in',

@@

In Eq. 8, \(P_{IM}(t)\) and \(P_{IF}(t)\) are the probabilities of
finding a male and female in `time in', respectively. There is no closed
form solution to the operational sex ratio, so we used recursion to
calculate \(\beta\) values for a given \(T_{\mathrm{f}}\),
\(T_{\mathrm{m}}\), and \(R\) (see Supporting Information S4),

@@

We formulate an individual-based simulation model to test whether the
predictions made by the analytical time-in time-out model are
qualitatively the same under a similar simulation model. We use the
individual-based model to test whether the prediction hold in finite
populations (see Supporting Information S6). The IBM was written in C.
All details of the initialisation, parameterisation and the specific
simulations which we run, are described in the supporting information
(see Supporting Information S1).

>>>>

The rates at which females are encountered by males
\(R_{\mathrm{f},\mathrm{m}}\) and males with nuptial gifts
\(R_{\mathrm{m_{G}}, \mathrm{f}}\) are encountered by females are both
calculated directly from the IBM, thereby modelling how these rates
might be estimated from empirical data, so the male threshold for
increasing fitness by searching is,

@@

Consequently, the IBM and the analytical model differ slightly (e.g.,
time is discrete in the IBM but continuous the analytical model, and in
the IBM, a fitness increment is applied to the focal female in the form
of birthrate increase rather than offspring fitness; see Supporting
Information S1 for details). But the predicted thresholds are
theoretically equivalent and yield predictions that are qualitatively
the same (Figure 3).

\begin{figure}
\centering
\includegraphics{ms_refs_fixed_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{The coevolution of male search and female choosiness as a
function of nuptial gift search time (\(\alpha\)). Points show where the
lower 95\% confidence interval of female choosiness (red) and male
search (blue) exceeds zero, indicating evolution of choosiness or
nuptial gift search. Each point includes data from 3200 replicate
simulations with identical starting conditions. Red and blue lines show
thresholds above which the mathematical model predicts that females
should be choosy and males should search, respectively (in agreement
with Figure 2). Minor deviations between the analytical model and the
simulation results are expected due to the finite population size, the
substantial stochasticity inherent to the simulation model, and the
potential for coevolution (for details, see Supporting Information S1).
The number of individuals in the population remained at or near carrying
capacity of \(K = 1000\). In each time step, up to 3000 total pair-wise
interactions occurred. Expected female processing time was set to
\(T_{\mathrm{f}}=2\) time steps, and \(\gamma\) and \(\alpha\) values in
the range {[}0.0, 2.0{]} and {[}0.1, 2.0{]}, respectively, were used.}
\end{figure}

Male thresholds \(\gamma_{\mathrm{m, IBM}}\) given by ineq. 4 accurately
predict the evolution of searching in the IBM across \(\alpha\) values,
and the female threshold \(\gamma_{\mathrm{f, IBM}}\) (ineq. 5) predicts
the evolution of female choice (Figure 3). In other words, IBM
simulations demonstrate that nuptial gift search in males, and
choosiness in females, will evolve from an ancestral state of no
searching and no choosiness in similar parameter space (Figure 3) as
predicted by the analytical model (Figure 2b). We further ran the IBM
with realistic values of \(\gamma\), estimated using data from the
\emph{P. mirabilis} system wherein choosiness among females, and nuptial
gift search among males, occur (Figure 4). We found that our IBM
predicts both the evolution of choosiness and nuptial gift searching
observed in the \emph{P. mirabilis} system.

\begin{figure}
\centering
\includegraphics{ms_refs_fixed_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{The joint evolution of male search and female choosiness using
a nuptial gift fitness increment (\(\gamma\)) that was estimated from
experimental data for the species \emph{Pisaura mirabilis} (mean
\(\gamma = 3.29 \pm\) SE via propagation of error for estimates of
nuptial feeding a non-nuptial feeding groups: {[}1.47; 8.32{]}). Points
show where the 95\% confidence interval exceeds 0 for female choosiness
(red) and male search (blue). Each point includes data from 100
replicate simulations with identical starting conditions. The red line
shows the threshold above which females should be choosy and the blue
line shows the threshold above which males should search. As predicted
by the analytical model, both male search and female choosiness evolved
for a range of \(\gamma\) values around the empirical estimates. This
occurs because these values result in a \(\gamma\) above the fitness
threshold necessary for selection to favour male search of nuptial gifts
(blue line) and female choosiness (red line). The number of individuals
in the population remained at or near carrying capacity of \(K = 1000\).
In each time step, up to 3000 total pair-wise interactions occurred.
Female processing time was set to \(T_{\mathrm{f}}=2\) time steps, and
\(\gamma\) and \(\alpha\) values in the {[}0.2, 2.0{]}, respectively,
were used. The parameter \(\gamma\) was estimated as relative increase
in offspring production such that \(\gamma\) is the factor by which
fitness increases (relative to the baseline fitness) given a nuptial
gift (See Supporting Information S5).}
\end{figure}

@@

Importantly, we show that the threshold nuptial gift value at which
females are favoured to express choosiness for nuptial gifts is rarely
equivalent to the threshold value at which males are favoured to search
for nuptial gifts, potentially leading to sexual conflict
(\protect\hyperlink{ref-Arnqvist2005a}{Arnqvist \& Rowe, 2005};
\protect\hyperlink{ref-Oliveira2008}{Oliveira \emph{et al.}, 2008}).
Here, we are defining sexual conflict as occurring when interactions
between sexes result in situation where both sexes cannot achieve an
optimal outcome simultaneously
(\protect\hyperlink{ref-Parker2006}{Parker, 2006}). As an example,
sizable areas of parameter space exists wherein the female optimum would
be to exhibit preference for (and receive) nuptial gifts, while the male
optimum is to not search for (and give) nuptial gifts (see Figure 2a,
Zone C). This will lead to mate encounters in which gift-less males will
benefit from mating, but females will not. In many systems, ecological
variables such as search time required to find a nuptial gift will
likely depend on prey abundance, which can vary substantially with time
in some species with nuptial gift-giving
(\protect\hyperlink{ref-Ghislandi2018}{Ghislandi \emph{et al.}, 2018}).
Since several ecological variables likely affect the value of these
thresholds, our results can be seen as providing some formalised
description of why nuptial gift-giving only occurs in some but not all
systems.

At first, the analytical model seems to suggests that nuptial gifts must
cause a very high fitness increase (approximately 25\%) before male
search and female choosiness is favoured by selection (Figure 2).
Similarly the IBM model seems to suggest that a fitness benefit of
approximately 50\% is required (see Figure 3). However, it is important
to recognise that these thresholds depend on multiple parameters. For
example, if female processing time (\(T_{\mathrm{f}}\)) is high, the
female threshold for choosiness with respect to \(\gamma\) drops such
that male search and female choosiness are favoured at lower \(\gamma\)
(see Supporting Information S2). If \(T_{\mathrm{f}}\) is sufficiently
high, then an initially rare gift-giving trait might be favoured by
selection even if the fitness benefit of a nuptial gift is low. The
effect that nuptial gifts have on fitness might vary across species, or
even populations. Effects on female fecundity have been estimated in
crickets, fireflies, butterflies, and spiders, but these estimates vary
considerably between species suggesting a large positive effect to no
effect at all (\protect\hyperlink{ref-Bergstrom2002}{Bergström \&
Wiklund, 2002}; \protect\hyperlink{ref-Rooney2002}{Rooney \& Lewis,
2002}; \protect\hyperlink{ref-Maxwell2018}{Maxwell \& Prokop, 2018};
\protect\hyperlink{ref-Gao2019}{Gao \emph{et al.}, 2019}).

We modelled the evolution of nuptial gift-giving using both a
mathematical model and an individual-based model. Our mathematical model
makes simple assumptions about the relationship between nuptial gift
search time (\(\alpha\)), conspecific encounter rate (\(R\)), female
processing time (\(T_{\mathrm{f}}\)), and the fitness increment of a
nuptial gift for offspring fitness (\(\gamma\)). It then derives the
threshold \(\gamma\) values above which males increase their fitness by
searching (\(T_{\mathrm{m}} > 0\)) for a nuptial gift
(\(\gamma_{\mathrm{m}}\)) and females increase their fitness by choosing
to reject males without gifts (\(\gamma_{\mathrm{f}}\)). In contrast,
our IBM models individuals over discrete time steps, and key processes
of nuptial gift acquisition, conspecific encounters, and female
processing are stochastic and varying among individuals. The
mathematical model and IBM make qualitatively identical predictions
(compare Figure 2b versus Figure 3), but differences between the two
models inevitably lead to quantitative differences for
\(\gamma_{\mathrm{m}}\) and \(\gamma_{\mathrm{f}}\) thresholds. For
example, \(\gamma_{\mathrm{f}}\) increased more rapidly with increasing
\(\alpha\) in the IBM compared to the analytical model. Some differences
are expected to occur due to stochastic effects inherent to IBMs (e.g.,
\protect\hyperlink{ref-Wilson2003}{Wilson \emph{et al.}, 2003}). Other
differences are more likely caused by more subtle assumptions between,
and limitations of, the two models. In general, the IBM did not do a
good job of controlling for conspecific interaction rate, making it
difficult to directly compare \(R\) between models. The IBM also allowed
for coevolution between male search and female choosiness (Figure 3),
which was not allowed in the analytical model (Figure 2). It was not our
goal to exactly recover the quantitative predictions of the analytical
model in our IBM. Future development of the IBM could further bridge the
gap between models while also developing new theory on how aspects of
the system such as explicit space, individual life history, or genetics
affect the evolution of nuptial gift-giving behaviour.

\hypertarget{nuptial-gift-giving-theory}{%
\subsection{Nuptial gift-giving
theory}\label{nuptial-gift-giving-theory}}

When modelling nuptial gift evolution, the challenge is to construct a
framework that captures the frequency-dependent selection between male
nuptial gift-giving and female choosiness for nuptial gifts, and we do
this using a time-in, time-out model. Recent studies have modelled some
frequency-dependent aspect of nuptial gift giving using evolutionary
game theory (\protect\hyperlink{ref-MaynardSmith1982}{Maynard Smith,
1982}; \protect\hyperlink{ref-Vincent2005}{Vincent \& Brown, 2005}). Two
such studies formulated a quantitative genetics model to study
evolutionarily stable nuptial gift sizes in populations where the female
propensity to re-mate was evolving
(\protect\hyperlink{ref-Kamimura2021}{Kamimura \emph{et al.}, 2021};
\protect\hyperlink{ref-Iwasa2022}{Iwasa \& Yamaguchi, 2022}). The
results obtained in these studies complement our results by giving
equilibrium solutions to the evolutionary stable nuptial gift size,
whereas we determine the general conditions under which nuptial
gift-giving will evolve as given by the inequalities we derive.

@@

\hypertarget{empirical-predictions}{%
\subsection{Empirical predictions}\label{empirical-predictions}}

@@

Our model assumes something akin to sequential polyandry. That is, a
system wherein female mating and reproduction with multiple males occurs
in sequence, rather than multiple matings occurring before reproduction.
In some systems with nuptial gift-giving, females have been documented
to mate multiple times before reproduction occurs, including the genus
of bark lice \emph{Neotrogla}
(\protect\hyperlink{ref-Kamimura2021}{Kamimura \emph{et al.}, 2021}),
and even our example system of \emph{P. mirabilis} where females will
sometimes engage in multiple mating before reproducing, especially if
starved because multiple mating may result in more nuptial gifts
(\protect\hyperlink{ref-Toft2015}{Toft \& Albo, 2015};
\protect\hyperlink{ref-Matzke2022}{Matzke \emph{et al.}, 2022}). It is
unclear what effect (if any) assuming non-sequential polyandry would
have on the threshold we derive. Under non-sequential polyandry, a
viable strategy for females might be to accept any male (with or without
gift) for fertilisation assurance, then exhibit a preference for nuptial
gifts. This might make choosiness less costly since it would entail less
of an opportunity cost to be choosy, and this could potentially make
female preference for nuptial gifts more likely to evolve. Our model
also assumes that males search in time-out, rather than contribute to
parental care, which is likely to be accurate for most systems but not
all. Expanding the model to explore these possibilities would be a
worthwhile goal for future research.

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

@@

Here we first present more detail for the derivation of fitness
threshold values \(\gamma_{\mathrm{m}}\) and \(\gamma_{\mathrm{f}}\). We
then present IBM simulations (see Supporting Information S1 for full
details). Code for simulations is available on GitHub (see ``Data
availability'').

@@

After mating, females must spend some time processing offspring
(\(T_{\mathrm{f}}\)). Male time to replenish sperm is assumed to be
negligible, but males can spend time out of the mating pool
(\(T_{\mathrm{m}}\)) to search for a nuptial gift. When males return
from time-out, they encounter females with some probability that is a
function of the rate at which an individual encounters conspecifics
(\(R\)) and the sex ratio (\(\beta\); males/females). Mortality occurs
for females and males in (\(\mu_{\mathrm{i,f}}\),
\(\mu_{\mathrm{i,m}}\)) and out (\(\mu_{\mathrm{o,f}}\),
\(\mu_{\mathrm{o,m}}\)) of the mating pool. Following Kokko \& Ots
(\protect\hyperlink{ref-Kokko2006}{2006}), we assume
\(m_{\mathrm{i,f}} = m_{\mathrm{o,f}} = 1\) and
\(m_{\mathrm{i,m}} = m_{\mathrm{o,m}} = 1\). While this choice is
arbitrary, we conducted a sensitivity analysis which shows that the
mortality parameters have no influence on the propensity for male search
and female choice to evolve (Supporting Information S2). First, we
describe the fitness consequences of male search time for a nuptial
gift. We then describe the fitness consequences of female choice to
accept or reject males based on their provision of a nuptial gift.

@@

During time-out, males have the opportunity to search for a nuptial
gift. Males can adopt one of two strategies; either search or do not
search for a nuptial gift. Males with the former strategy continue to
search until they find a nuptial gift, while males that do not search
will immediately re-enter the mating pool. In this case, time searching
for a nuptial gift will come at the cost of mating opportunities but
might increase offspring fitness. We therefore need to model the
expected length of time \(E[T_{\mathrm{m}}]\) spent outside of the
mating pool for males that search for nuptial gifts, which is simply
\(\alpha\). Note that we can integrate search time \(t\) over the rate
at which nuptial gifts are encountered (\(\exp(-1/\alpha)\)) to show
\(E[T_{\mathrm{m}}] = \alpha\),

@@

The rate at which a focal male that searches for a nuptial gift
(\(\mathrm{M_{G}}\)) increases his fitness is therefore the fitness of
his offspring \((1 + \gamma)\) divided by expected time spent searching
for a nuptial gift (\(\alpha\)) plus time spent in the mating pool,
\((\beta + 1)/R\) (recall that females produce \(\lambda\) offspring),

\[W_{\mathrm{M_{G}}} = \lambda \frac{1 + \gamma}{\alpha + \left( \frac{\beta + 1}{R} \right)}.\]

In contrast, a male that does not search for a nuptial gift
(\(\mathrm{M_{L}}\)) has offspring with lower fitness, but spends less
time outside of the mating pool,

\[W_{\mathrm{M_{L}}} = \lambda \frac{1}{\left(\frac{\beta+1}{R} \right)} = \lambda \frac{R}{\beta + 1}.\]

We can then determine the conditions for which
\(W_{\mathrm{M_{G}}} > W_{\mathrm{M_{L}}}\), isolating \(\gamma\) to
find how large of a fitness benefit must be provided by the nuptial gift
to make the search cost worthwhile, which simplifies to ineq. 2. When
this inequality holds, males are favoured to search until they find a
nuptial gift, which would result in an average search time of
\(\alpha\). When the male trait is continuous (i.e., males search for
time period \(T_{\mathrm{m}}\)), it can be shown that the same threshold
can be reached by evaluating the partial derivative of the male fitness
function (Supporting Information S3). Hence, the thresholds are
consistent under different assumptions concerning male searching
strategy. Selection will cause males to search for nuptial gifts if the
fitness increase to offspring exceeds the product of search time and
female encounter rate.

@@

\[R_{\mathrm{F_{G}}} = R \left(\frac{\beta}{\beta + 1}\right)\left(1 - e^{-\frac{1}{\alpha}T_{m}}\right).\]

@@

\[R_{\mathrm{F_{L}}} = R \left(\frac{\beta}{\beta + 1}\right)\left(e^{-\frac{1}{\alpha}T_{m}}\right).\]

@@

If \(R_{\mathrm{F_{G}}}\) is sufficiently high and
\(R_{\mathrm{F_{L}}}\) is sufficiently low, then finding a male with a
gift will be easier than finding a male without one. Also, the expected
time spent in the mating pool before a focal female encounters a male
with a gift will be \(1/R_{\mathrm{F_{G}}}\), while the expected time
spent in the mating pool before a focal female encounters any male will
be \(1 / (R_{\mathrm{F_{G}}} + R_{\mathrm{F_{L}}})\). Finally, the rates
at which a female encounters males with and without a gift,
\(R_{\mathrm{F_{G}}}\) and \(R_{\mathrm{F_{L}}}\), are different from
the probabilities that a male encounter has or does not have a gift. The
rate of encounter is no longer relevant in this case because we are
assuming that an encounter has occurred. Hence, the probability of an
encountered male having a gift is simply,

\[P(G) = \frac{1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}}{\left(1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}\right) + e^{-\frac{1}{\alpha}T_{\mathrm{m}}}} = 1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@

\[P(L) = e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.\]

@@

\begin{equation}
W_{\mathrm{F_{G}}} = \lambda \frac{1 + \gamma}{T_{\mathrm{f}}} + \frac{1}{\mathrm{F_{G}}}.
\end{equation}

The top of the right-hand side of Eq. 6 gives the fitness increase, and
the bottom gives the total time it takes to obtain this fitness. The
\(R_{\mathrm{F_{G}}}\) is inverted because it represents the expected
time to encountering a male with a gift. We can expand Eq. 6,

\[W_{\mathrm{F_{G}}} = \lambda \frac{1 + \gamma}{T_{\mathrm{f}} + \frac{1}{R \left(\frac{\beta}{\beta + 1}\right)\left(1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}\right)}}.\]

@@

\[W_{\mathrm{F_{G,L}}} = \lambda \frac{\left(1 + \gamma\right)\left(1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}\right) + e^{-\frac{1}{\alpha}T_{\mathrm{m}}}}{T_{f} + \frac{1}{R \left(\frac{\beta}{\beta + 1}\right)}}.\]

We then evaluate the conditions under which
\(W_{\mathrm{F_{G}}} > W_{\mathrm{F_{G,L}}}\). We isolate \(\gamma\) to
determine how much offspring fitness must be increased to make
choosiness beneficial (\(\gamma_{\mathrm{f}}\)),

@@

We assume that the sex ratio is equal upon maturation. Given this, Kokko
\& Monaghan (\protect\hyperlink{ref-Kokko2001}{2001}) show that the
operational sex ratio depends on the probability of finding an
individual in time-in,

@@

In Eq. 8, \(P_{IM}(t)\) and \(P_{IF}(t)\) are the probabilities of
finding a male and female in time-in, respectively. There is no closed
form solution to the operational sex ratio, so we used recursion to
calculate \(\beta\) values for a given \(T_{\mathrm{f}}\),
\(T_{\mathrm{m}}\), and \(R\) (see Supporting Information S4),

@@

We formulate an individual-based simulation model to test whether the
predictions made by the analytical time-in time-out model are
qualitatively the same under a similar simulation model. We use the
individual-based model to test whether the prediction hold in finite
populations. The IBM was written in C. All details of the
initialisation, parameterisation and the specific simulations which we
run, are described in Supporting Information S1.
